---
title: "MTH6139 Time Series" 
subtitle: "Coursework 1 - Apple Stock Analysis" 
author: "Varunika Chellarajesh" 
date: "15.03.2026" 
output: 
  html_document:
    toc: true
    toc_float: true
    theme: spacelab 
    highlight: tango
---
 
```{r, echo=FALSE}
# This code will display the QMUL logo at the top right of the page
# Do not change this code
htmltools::img(src = knitr::image_uri("images/QMlogo.png"),
               alt = 'logo',
               style = 'position:absolute; top:0; right:0; padding:10px; width:20%;')
```

# Section 1: Prophet Forecasting for Apple Stock

## Introduction

This project applies Meta’s Prophet forecasting system to Apple stock prices.

The dataset contains daily observations of Apple stock prices (ticker symbol: AAPL), taken from Kaggle.

Events like  COVID-19, global conflicts, and changes in the technology sector have had significant impacts on stock price.

The aim is to understand how the series behaves and to produce forecasts.

---

## Data Import

The data were imported from a CSV file stored in the `data` folder.

```{r data-import}
apple_data_raw <- read.csv("data/apple.csv")
```


## Data Preparation for Prophet

The date column was converted to date format and renamed ds. The stock prices were cleaned and stored in a variable named y.

```{r}
apple_prophet_data <- data.frame(
  ds = as.Date(apple_data_raw$Date, format = "%m/%d/%Y"),
  y = as.numeric(gsub("[\\$,]", "", apple_data_raw$Close.Last))
)

apple_prophet_data <- na.omit(apple_prophet_data)
```

## Visualisation of the Time Series

Before fitting the model, it is useful to examine the raw time series.

```{r}
plot(
  apple_prophet_data$ds,
  apple_prophet_data$y,
  type = "l",
  main = "Apple Stock Price",
  xlab = "Date",
  ylab = "Price"
)
```

The series shows a general upward trend over time, with noticeable short-term fluctuations.

## A Simple Linear Trend Check

This confirms the upward trend seen in the data.

```{r}
time_index <- 1:nrow(apple_prophet_data)

linear_trend_model <- lm(y ~ time_index, data = apple_prophet_data)

summary(linear_trend_model)
```

The output gives a  summary of the overall direction of the series.


## Fitting the Prophet Model

The Prophet package was used to fit a forecasting model to the prepared Apple stock data.

```{r}
library(prophet)

apple_prophet_model <- prophet(apple_prophet_data)
```

Prophet models the series using trend and seasonal components.

## Creating Future Dates

Future dates were generated in order to produce forecasts.

```{r}
apple_future_dates <- make_future_dataframe(apple_prophet_model, periods = 8)
```

## Generating Forecasts

The fitted model was used to generate forecasts.

```{r}
apple_forecast <- predict(apple_prophet_model, apple_future_dates)
```


## Forecast Output

It is useful to inspect some of the forecast values.

```{r}
tail(apple_forecast[, c("ds", "yhat", "yhat_lower", "yhat_upper")])
```

The table shows predicted values together with uncertainty intervals.

## Prophet Forecast Plot

The main forecast plot is shown below.

```{r}
plot(apple_prophet_model, apple_forecast)
```

The black points show the observed data, while the blue line shows the fitted and forecast values.

## Prophet Components

The component plots help to separate the main parts of the Prophet model.

```{r}
prophet_plot_components(apple_prophet_model, apple_forecast)
```

These plots show how Prophet decomposes the time series.

## A Shorter Time Frame

It is useful to look at a more recent part of the data.


```{r}
recent_apple_data <- subset(
  apple_prophet_data,
  ds >= as.Date("2020-01-01")
)

plot(
  recent_apple_data$ds,
  recent_apple_data$y,
  type = "l",
  main = "Apple Stock Price Since 2020",
  xlab = "Date",
  ylab = "Price"
)
```

Looking at a shorter time frame makes it easier to observe recent movements in the series.

## Comments on the Results

Overall, Apple stock prices show a clear upward trend over time, although there are some fluctuations along the way. 

There is a noticeable drop around early 2020, which is likely due to the COVID-19 pandemic, when financial markets were very uncertain.

After this period, the stock recovers and continues to grow, possibly because of increased demand for technology and digital services.

There are also some ups and downs in the later years, which could be linked to global events and changes in the economy.

In particular, factors such as interest rates and general market conditions seem to affect the stock price at different points in time.

Overall, the stock has grown over time, but it is still influenced by many outside factors.

## Conclusion

In this report, Apple stock price data were analysed using the Prophet model.

The results show that Prophet is easy to use and produces clear visual outputs. It provides a simple way to understand trends and generate forecasts for time series data.
